---
title: 'A script for extracting summary and descriptive statistics from FoodMicrobionet METAOlive subset (v5.0.1)'
subtitle: "Bacteria"
author: "Prof. E. Parente, DAFE, Università degli Studi della Basilicata"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
---

**NOTES FOR SELF**

* What do we do for olives to which spices/herbs are added? Technically one may add a comma delimited vector of spices but data wrangling may become complicated; however, this information might be important; plus, what do we use to identify spices? Common names? Latin names? Something with a stronger link with FoodOn or FoodEx?  

* I have left for later the (optional) rarefaction and rarefaction analysis using Vegan functions, although there is a phyloseq function as well: rarefy_even_depth()  

* in next iteration remove the ad hoc statements (I will fix everything inside FMBN)

* this version uses FMBN v 5.0.1: some problems with sample metadata which required ad hoc fixes here were fixed in the sample table of FMBN and the ad hoc code will not be needed


```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

# loading packages
.cran_packages <- c("tidyverse", "randomcoloR", "knitr", "rnaturalearth", "ggrepel",
                    "rnaturalearthdata", "readxl", "beepr", "vegan", "crayon",
                    "kableExtra")
.bioc_packages <- c("BiocManager", "phyloseq")

# package PERFect is not needed with filter_my_taxa v 1.2 onward

.inst <- .bioc_packages %in% installed.packages()
if(any(!.inst)) {
  if(!.inst[1]) {
    install.packages("BiocManager")
    .inst <- .bioc_packages %in% installed.packages()
  }
  if(any(!.inst[2:length(.inst)])) {
    BiocManager::install(.bioc_packages[!.inst], ask = F)
  }
}

.inst <- .cran_packages %in% installed.packages()
if(any(!.inst)) {
  install.packages(.cran_packages[!.inst])
}
sapply(c(.cran_packages,.bioc_packages ), require, 
       character.only = TRUE)

opar <- par(no.readonly=TRUE) 
par(ask=F) 
set.seed(1234) 
world <- ne_countries(scale = "medium", returnclass = "sf")

```

```{r options_functions, echo = F, include=F}

## load functions
fn_path <- "source"
# load the filter_my_taxa function
source(file.path(fn_path,"filter_my_taxa_function_1_2.R"))
# open and read the function for more details

# load the make_OTU_palette function
source(file.path(fn_path,"make_OTU_palettes.R"))

# load the prevab_pc function (needed for ordering prevalence and abundance tables)
source(file.path(fn_path,"pc_prevab.R"))

# extra options
verbose_output <- T # this prints extra output (tables mostly). 
# Set to F if you don't want it.

play_sound <- T # plays a sound with beepr::beep() when execution is completed
which_sound <- 6 # the sound you are going to hear

outfn_prefix <- "FMBN_METAO_B_0225" # a prefix used for saved output files
out_folder <- "output_bacteria" # the output folder
if(!dir.exists(out_folder)) dir.create(out_folder)

# options for saving graphs
savegraph <- T  # set savegraph to T if you want to save graphs 
# (some graphs won't be displayed, only saved); use F if you don't want to save
graphresolution <- 300 # in dpi
graphsizein <- 7 # size in inches, overridden in some graphs
graphtype <- "tif" # alternative value is "pdf", avoid using "jpg"
gfileext <- ifelse (graphtype == "pdf", ".pdf", ".tiff") 

#digits in tables
dgts <- 3

# options for number of OTU groups in sample and OTU palettes
overrides <- F # set to T if you want more than 15 categories in sample palettes
overrideOTU <- F # set to T if you want more than 15 categories in OTU palettes

# The "Other" column has no taxonomic meaning unless it is an aggregate of
# several OTUs (see instructions in the FMBN_make_files* script) and will be
# removed by default. The "Chloroplast" and "Mitochondria" columns only indicate  
# contamination with chloroplast or mitochondria  DNA/RNA and should also be  
# removed. Set to F if you want to keep these columns in OTU tables.
# Eukaryota can also be removed
removeOther <- T
removeKonly <- T # removes the taxa which have been identified only at the level of Kingdom
removeChloroplast <- T # only needed if you work on 16S as a target
removeMitochondria <- T # only needed if you work on 16S as a target
removeFungi <- T # F if you are working on fungi, T if you are working on bacteria

glom_level <- "genus"
# may be used in tax_glom; will remove anything which has an identification above
# this level and agglomerate the taxa at the tax_glom level; use wisely
# "none" no agglomeration, "genus", "family", "order", "class"
# will return an error if not in rank_names()

# options for filtering samples and OTUs/features

# if filtersamples == T the sample filter will be applied (overridden if sample_agg == "exp. code")
# if filterOTUs == T a prevab or permutation filter will be applied to OTUs/features
filtersamples <- T
# samples/sample groups with less than minsampleab sequences will be removed
minsampleab <- 1000 

filterOTUs <- T
prev_filter <- "prevab" # the type of filter, prevab or PERFect
# filter for OTU abundance. OTUs with a maximum relative abundance < abthreshold
# will be removed or pooled (see below)
# keep the value low or you will loose too many OTUs between filtering and
# rarefaction
ab_threshold <- 0.005
# options for taxa prevalence filter 
prev_threshold <- 0.01 # as fraction of samples
prev_filter_flag <- F
pass_both <- T # both the prevalence and abundance filter must be passed

# sorting the prevalence and abundance table, this is only important for selecting top OTUs
prevab_sort_option <- "comp1" 
# alternatives are
# decr_prev: by decreasing prevalence, the default; 
# decr_ab: by decreasing average relative abundance, regardless of the prevalence
# prevab: decreasing prevab, 100*relAbundance*relprev); not very advisable
# comp1: uses the results of the prevab_pc function; a reasonable compromise

# flags for filtering, will be set to true if filtering is performed
filtersamples_flag <- F
filterOTU_flag <- F

# NOTE FOR SELF: need to fix this is I decide to rarefy
# rarefy? if rareoption<-T vegan::rrarefy() will be used to extract a random rarefied
# matrix. If raremin is T will rarefy to the closest hundred below the
# minimum number of sequences, otherwise will rarefy to raren.
# The default is F, although rarefaction will be performed for bipartite
# analysis 

rareoption <- T
raremin <- T
raren <- 10000
rarefy_flag <- F # will be set to T if rarefaction is performed

# pooling of filtered OTUs: if pool is T, abundances of OTUs removed by 
# filtering are pooled and shown as "Other" in bar charts; does not apply
# to heat maps and MDS.
pool <- T 

# maximum number of OTUs allowed in stacked bar charts
# will be used to generate the OTU palette with randomcoloR
maxOTUcats <- 25
# maximum number of "species" for which prevalence and abundance is saved
topp <- 50
# if the number of "species" in data is lower, then it is going to be the 
# max number of species
# maximum number of "species" in box plots, heat maps and NMDS
# must be <topp
topn <-25
# max "samples" or sample categories for barplots and boxplots
max_samples <- 25
# max number of samples in the bonus barplot (the last chunk!)
topx <- 19

# logging the color scale in heat maps; set to T if you want the scale
# in log10(relabundance); set to F if you want to keep a linear scale
loghmscale <- T
# option for performing rarefaction analysis on the unfiltered matrix
dorareanalysis <- F



```


```{r loading_data, echo = F, message = F}

metaolive_phy_B <- readRDS(file.path("data", "olive_FMBN_5_B_phy_mod.rds"))
# extract components 
sample_data_B <- as(sample_data(metaolive_phy_B), "data.frame")
taxa_B <- as(tax_table(metaolive_phy_B), "matrix")
taxa_B <- as(tax_table(metaolive_phy_B), "matrix")
OTU_B <- as(otu_table(metaolive_phy_B), "matrix")

# get studies
studies <- read_xlsx(file.path("data", "FMBN_5_olive_studies.xlsx"))

# filter and retain only the ones with bacteria
studies_B_vector <- pull(sample_data_B, studyId) |> unique()

studies_B <- studies |>
  dplyr::filter(studyId %in% studies_B_vector)

# get extra metadata 
extra_metadata_B <- read_xlsx(file.path("data", "extra_metadata_FMBN_B.xlsx"), sheet = "Metadata")

# join the metadata, plus add a few other grouping or labeling variables
sample_data_B_extra <- sample_data_B |> 
  rownames_to_column("SRA_run") |>
  rename(spoil_ferm = spoilage)
# add further variables from metadata
sample_data_B_extra <- left_join(sample_data_B_extra,
                                 select(extra_metadata_B, sampleId, descr = description, DOI,
                                        olive_cultivar:Duration_of_storage_post_fermentation_d)
                                 )
# a few mutate ops
sample_data_B_extra <- sample_data_B_extra |>
  mutate(olive_cv_short = str_sub(olive_cultivar, start = 1, end = 3),
         olive_var_short = str_sub(olive_variety, start = 1, end = 3),
         sample_type_short = case_when(
           sample_type == "fruit" ~ "F",
           sample_type == "brine" ~ "B",
           sample_type == "fruit+brine" ~ "BF",
           .default = "O"
         )
  ) |>
  mutate(group_short = str_c(olive_var_short, sample_type_short, sep ="_")) |>
  mutate(nature = if_else(is.na(nature), "Other", nature),
         process = if_else(is.na(process), "Other", process),
         spoilage = if_else(is.na(spoilage), "Other", spoilage)) |>
  mutate(ripe_trade = str_c(str_remove(sample_data_B_extra$olive_ripeness, " olives"),
                            str_remove(sample_data_B_extra$olive_trade_prep, " olives"), sep = " ")) |>
  column_to_rownames(var = "SRA_run")

# write_tsv(sample_data_B_extra, file = file.path("data", "sample_B_extra_0125.tsv"))
# put the metadata back into the phyloseq
sample_data(metaolive_phy_B) <- sample_data_B_extra

# any further operation will be performed on a copy
myphyseq <- metaolive_phy_B
saveRDS(myphyseq, file.path("data", "olive_FMBN_5_0_1_B_extra_phy.rds"))

# quick check
# View(as(sample_data(metaolive_phy_B), "data.frame"))
```

# FMBN-METAOlive statistics.  

This report is designed to provide summary statistics for the subset of data on table olives extracted from FoodMicrobionet, version 5.0 or higher. It takes as an input a merged phyloseq object with extra metadata and returns summary data and statistics and a few descriptive graphs. You can personalize most of your analysis in the **options_functions** chunk.    


# Statistics on studies.

```{r get_descr_stats, echo = F, message = F}
n_studies <- nrow(studies)
n_studies_16S_only <- nrow(dplyr::filter(studies_B, !str_detect(target, "ITS")))
n_studies_ITS_only <- nrow(dplyr::filter(studies, str_detect(target, "ITS") & !str_detect(target, "16S")))  
n_studies_both <- nrow(dplyr::filter(studies, str_detect(target, "ITS") & str_detect(target, "16S")))
min_samples_study <- min(studies$samples)
max_samples_study <- max(studies$samples)
median_samples_study <- median(studies$samples)
```


The number of studies in this version of METAOlive FMBN is `r n_studies`. This version includes studies on bacterial microbiota only (`r n_studies_16S_only`), on fungal microbiota only (`r n_studies_ITS_only`) and studies for which both data for bacterial and fungal microbiota are available (`r n_studies_both`). However, due to inconsistencies in the deposit of sequences in SRA (in several cases the same sample was deposited with two separate biosample accessions and/or data for bacteria and fungi were deposited with different bioproject of study accessions), the same samples might be present in two studies, one for bacteria and one for fungi^[by convention in FoodMicrobionet a study must have a unique bioproject accession and a sample a unique biosample accession]. We did our best to match samples in these situations. The addition of datasets on fungi is in progress and, when available, we will progressively add fungal data for all the studies which are already in FoodMicrobionet with bacterial community data and add more fungal studies.  
There are **`r nrow(sample_data_B)`** samples on bacterial communities of table olives in this version.  
The experimental approaches vary and both longitudinal and cross-sectional studies are available. The number of samples per study ranges from `r min(studies_B$samples)` to `r max(studies_B$samples)`, with a medin of `r median(studies_B$samples)` samples per study.  

# FMBN-METAOlive statistics for studies on bacteria. 

## Platforms, gene target, regions.  

FMBN grows by addition of sequences deposited in NCBI SRA for published studies. As a consequence, use of targets (16S RNA, 16S RNA gene, ITS) reflect what is published and the correlation between platforms and targets.  

```{r platform, dpi = 96, echo = F}

platform_region <- dplyr::select(studies_B, studyId, ref_short, platform, target, region, samples, year)

platform_region <- platform_region %>%
  mutate(
    platform_2 = case_when(
      str_detect(platform, "454 GS") ~ "454 GS",
      str_detect(platform, "Illumina") ~ "Illumina",
      str_detect(platform, "Ion Torrent") ~ "Ion Torrent"
    )
  ) %>%
  mutate(year = as.numeric(year)) %>%
  arrange(year)
kable(select(platform_region, -platform), caption = "Table 1. Platforms and targets. Studies including data on bacteria in table olives") |>
  kableExtra::kable_classic()

```

All recent studies use Illumina as a platform and V3-V4 region of the 16S RNA gene as a target. Data on fungi will be shown in a separate report.  


# Statistics on samples.  

The number of samples per study ranges from `r min(studies_B$samples)` to `r max(studies_B$samples)`, with a median of `r median(studies_B$samples)` samples per study. 
The geographic distribution of samples is shown below.

```{r samples_by_location, echo = F, message = F, warning=F}
# creating a table for continents
table_by_continent <- sample_data_B_extra %>%
  group_by(geo_loc_continent) %>%
  count(geo_loc_continent) %>%
  ungroup() %>%
  arrange(desc(n)) %>%
  mutate(prop=n/sum(n)) %>%
  mutate(cumprop = cumsum(prop))
kable(table_by_continent, digits = c(0,4,4,4)) |> kable_classic()
# creating a table for countries
table_by_country <- sample_data_B_extra %>%
  group_by(geo_loc_country) %>%
  count(geo_loc_country) %>%
  ungroup() %>%
  arrange(desc(n)) %>%
  mutate(prop=n/sum(n)) %>%
  mutate(cumprop = cumsum(prop),
         geo_loc_country = if_else(geo_loc_country == "USA", "United States of America", geo_loc_country))
kable(table_by_country, digits = c(0,rep(dgts,3)))

my_countries <- sample_data_B_extra$geo_loc_country |> unique()

world_2 <- left_join(world, 
                     dplyr::select(table_by_country, sovereignt=geo_loc_country, n_samples = n)) %>%
  mutate(logn_samples = log10(n_samples)) %>%
  dplyr::filter(sovereignt %in% my_countries & str_detect(type, "Country|country"))

# ggplot(data = world_2) +
#  geom_sf(aes(fill = logn_samples)) +
# scale_fill_viridis_c(direction = -1, na.value = "white") +
#  labs(fill = "log10(samples)",
#       title = "Number of samples, by country") +
#  theme_bw() +
#  theme(plot.title = element_text(hjust = 0.5))
```
```{r plot_map, echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, results='hide',fig.keep='all'}
ggplot(data = world_2) +
  geom_sf(aes(fill = n_samples), size = I(0.2)) +
 scale_fill_viridis_c(direction = -1, na.value = "white", trans = "log2") +
  labs(fill = "samples (log2 scale)",
       title = "Number of samples, by country, data on bacterial communities") +
  coord_sf(xlim=c(-10, 40), ylim=c(30, 55), expand = F)
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))


# ggsave(filename = "nsamples_by_country.tiff", dpi = 300, units = "in", width = 9, height = 5)

if(savegraph)  ggsave(filename = file.path(out_folder,"nsamples_by_country.jpg"), dpi = 300)

```

Samples were available for fruits, brine,  or contact surface and raw materials.
 
```{r by_sample_type, echo = F}
sample_data_B_extra |> 
  mutate(as.factor(sample_type)) |>
  count(olive_trade_prep, sample_type)
```

`r nrow(dplyr::filter(sample_data_B, target1 == "16S_RNA"))` samples out of `r nrow(sample_data_B)`have 16S RNA as a target, and will be removed before processing.  

```{r filter_by_target, echo = F}
# nsamples(myphyseq)
myphyseq <- subset_samples(myphyseq, target1 != "16S_RNA")
# nsamples(myphyseq)
```

## Statistics by olive variety.  

The data were extensively annotated using information from NCBI sample metadata files and information from the publications. Metadata included olive tree cultivar, olive ripeness, olive process, style, occurrence of spoilage, use of starters, duration of fermentation and of storage post fermentation (if known). A summary table is available below.     

```{r summary_cultivar}
sample_summary_pre_filter <- sample_data_B_extra |> 
  count(olive_trade_prep, olive_cultivar, olive_variety, olive_var_short, olive_ripeness, sample_type)
kable(sample_summary_pre_filter, caption = "Sample summary, before sample filtering") |> kable_classic()
out_fn <- file.path(out_folder, str_c(outfn_prefix,"samples_prefilter.txt"))
write_tsv(sample_summary_pre_filter, file = out_fn)
```


# The bacterial microbiota of table olives.  

I will now use a short pipeline for data analysis with the following steps:  

1. filtering samples: samples with less than `r minsampleab`

3. perform diversity estimation (I have left an option for rarefaction)

2. filtering and agglomeration of OTU/features will perform the following operations:

  + perform agglomeration with phyloseq::tax_glom() if applicable  
  
  + remove poorly identified taxa (skipped if tax_glom is performed)  
  

## Sample filtering.  
  
```{r sample_filt, echo = F}
# add any additional filter you want here
n_samples_prefilter <- nsamples(myphyseq)
if(filtersamples){
  myphyseq <- prune_samples(sample_sums(myphyseq)>=minsampleab, myphyseq)
  filtersamples_flag <- T
}

sample_summary_post_filter <- sample_data_B_extra |> 
  dplyr::filter(n_reads2>=minsampleab) |>
  count(olive_trade_prep, olive_cultivar, olive_variety, olive_var_short, olive_ripeness, sample_type)
kable(sample_summary_pre_filter, caption = "Sample summary, after sample filtering") |> kable_classic()
out_fn <- file.path(out_folder, str_c(outfn_prefix,"samples_postfilter.txt"))
write_tsv(sample_summary_pre_filter, file = out_fn)

n_samples_postfilter <- nsamples(myphyseq)
```
There were `r n_samples_prefilter` in the original phyloseq object. After removing samples with less than `r minsampleab` reads, there are `r n_samples_postfilter` samples left.  



## Alpha diversity.  

Here I will compare the alpha diversity (in a few reasonable comparisons) before any taxa filtering or agglomeration.

```{r alpha_diversity, echo = F, dpi = 96}

myphyseq %>% plot_richness(x = "olive_var_short", shape = "sample_type",
              color = "nature",
              measures = c("Shannon", "Chao1")) +
  scale_shape_manual(values=c(16,17,15,18)) +
  theme_bw()

alpha_div <- estimate_richness(myphyseq, measures = c("Shannon", "Chao1"))

# to do: join metadata and plot using ggplot2
# may be use facets, fills for starter, alpha for duration, need to include olive_trade prep
# may be add jitter; consider using free for facets
sample_data_B_2 <- sample_data_B_extra |> 
  rownames_to_column("Run")

# ad hoc fixes to be removed from version 5.1 of FoodMicrobionet (changes were made in FMBN sample table)

rows_to_change <- which(sample_data_B_2$sampleId %in% c(504,515) )
sample_data_B_2$nature[rows_to_change]<- "Finished"
sample_data_B_2$process[rows_to_change]<- "None"
sample_data_B_2$spoilage[rows_to_change]<- "Fermented"
rows_to_change <- which(sample_data_B_2$sampleId %in% c(505,506,509,512,513,514, 518,519,520,521) )
sample_data_B_2$nature[rows_to_change]<- "Intermediate"
sample_data_B_2$process[rows_to_change]<- "None"
sample_data_B_2$spoilage[rows_to_change]<- "Fermented"
alpha_div_wmetadata <- alpha_div |>
  rownames_to_column("Run") |>
  left_join(sample_data_B_2)
min_Chao1 = min(alpha_div_wmetadata$Chao1)
max_Chao1 = max(alpha_div_wmetadata$Chao1)
  
alpha_div_wmetadata |>
  ggplot(mapping = aes(x = olive_var_short, y = Chao1)) +
  geom_jitter(mapping=aes(shape = sample_type, color = olive_trade_prep, 
                          alpha = Duration_of_fermentation_d)) +
  scale_shape_manual(values=c(16,17,15,18)) +
  scale_alpha_continuous(range = c (0.3,1)) +
  theme_bw()

# as an alternative create a new variable and use categories

alpha_div_bacteria <- alpha_div_wmetadata |>
  mutate(stage_category = case_when(
    nature == "Raw" ~ 1,
    nature == "Intermediate" ~ 2,
    nature == "Finished" ~ 3,
    .default = 0
  ))

alpha_div_bacteria_plot <- alpha_div_bacteria |>
  ggplot(mapping = aes(x = olive_var_short, y = Chao1)) +
  geom_boxplot() +
  geom_jitter(mapping=aes(shape = sample_type, color = ripe_trade,
                          size = stage_category), alpha = I(0.5)) +
  labs(
    x = "olive variety",
    color = "olive ripe. + trade prep.",
    shape = "sample type",
    size = "stage"
  ) +
  scale_shape_manual(values=c(1,2,0,5)) +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  theme_bw()
alpha_div_bacteria_plot
out_fn <- file.path(out_folder, str_c(outfn_prefix,"_alphadivbacteria",gfileext)) 

if(savegraph) ggsave(alpha_div_bacteria_plot, filename = out_fn, dpi = graphresolution)



# is there a relationship between number of sequences and diversity?
alpha_div_bacteria |>
  ggplot(mapping = aes(x = n_reads2, y = Chao1)) +
  geom_point()
summary(lm(Chao1~n_reads2, data = alpha_div_wmetadata))
# not really

# is there a relationship with addition of starters?
# first, which studies have a combination of starter and no starter?
# xtabs(~studyId+inoculation_fortification, data = alpha_div_wmetadata)
# ST133, ST229, ST253
alpha_starter <- alpha_div_bacteria |> 
  dplyr::filter(studyId %in% c("ST133", "ST229", "ST253")) |>
  ggplot(mapping = aes(x=inoculation_fortification, Chao1)) +
  facet_wrap(~studyId) +
  geom_boxplot() +
  geom_jitter(mapping=aes(shape = sample_type, color = ripe_trade,
                          size = stage_category), alpha = I(0.5)) +
  labs(
    x = "inoculation",
    color = "olive ripe. + trade prep.",
    shape = "sample type",
    size = "stage"
  ) +
  scale_shape_manual(values=c(1,2,0,5)) +
  scale_size(breaks = c(1,2,3)) +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  theme_bw()
alpha_starter
out_fn <- file.path(out_folder, str_c(outfn_prefix,"_alphadivbacteria_starter",gfileext)) 

if(savegraph) ggsave(alpha_starter, filename = out_fn, dpi = graphresolution)

# any evidence of variation with stage?
alpha_div_stage <- alpha_div_bacteria |>
  mutate(stage_category = as.factor(stage_category)) |>
  ggplot(mapping = aes(x=stage_category, Chao1)) +
  facet_wrap(~olive_variety) +
  geom_boxplot() +
  geom_jitter(mapping=aes(shape = sample_type, color = ripe_trade), alpha = I(0.5)) +
  labs(
    x = "stage",
    color = "olive ripe. + trade prep.",
    shape = "sample type"
  ) +
  scale_shape_manual(values=c(1,2,0,5)) +
  scale_size(breaks = c(1,2,3)) +
  scale_color_brewer(type = "qual", palette = "Dark2") +
  theme_bw()
alpha_div_stage
out_fn <- file.path(out_folder, str_c(outfn_prefix,"_alphadivbacteria_stage",gfileext)) 

if(savegraph) ggsave(alpha_div_stage, filename = out_fn, dpi = graphresolution)


```
Due to the differences in methods and experimental designs only very general conclusions can be drawn:

* there is no evidence of a relationship between number of sequences and Chao1

* olive variety and process are almost always confounded in the same study, therefore comparisons need to be done, at least, within the same variety  

* for the few studies for which inoculated and inoculated olives were compared, we found no significant evidence of lower diversity in inoculated fermentations; median Chao1 was even higher in some cases for inoculated fermentations 

* there is no clear evidence of stage affecting the diversity, i.e. diversity does not apparently decrease with increasing fermentation time due to the selection of a less diverse and more adapted community, although in some environmental samples (see Nyons olives) diversity is much higher than in olives or brines.  

```{r rarefy, echo = F, include = F}
# to do, a chunk on rarefaction
```

## Filtering taxa: taxonomic agglomeration, prevalence and abundance.  

I will now use a prevalence and abundance filer (a permutation filter would be more useful) and create prevalence and abundance graphs and tables. Please note that top taxa will be selected in decreasing order of importance according to one of three possible criteria: prevalence, and index obtained by multiplying prevalence and abundance, and the scores of the first component of the relative prevalence and mean abundance matrix. The current option is `r prevab_sort_option`.

```{r prev_ab, echo = F}

ntaxa_pre <- ntaxa(myphyseq)

# keep sample sums (needed to keep track of what is happening)
sample_sums_stage_0 <- sample_sums(myphyseq)

# first filtering stage
myphyseq_1 <- myphyseq
# apply taxonomic filters or agglomerate
# optionally remove fungi or Eukaryota
if (removeOther) {
  myphyseq_1 <- subset_taxa(myphyseq_1, taxa_names(myphyseq_1)!="Other")
}
if (removeChloroplast){
  myphyseq_1 <- subset_taxa(myphyseq_1, 
                            taxa_names(myphyseq_1)!="Chloroplast" & class!= "Chloroplast") 
}
if (removeMitochondria){
  myphyseq_1 <- subset_taxa(myphyseq_1, 
                            taxa_names(myphyseq_1)!="Mitochondria" & family != "Mitochondria")
}
if (removeFungi){
  myphyseq_1 <- subset_taxa(myphyseq_1, 
                            domain !="Mitochondria" & domain != "Fungi")
}

ntaxa_post_stage1 <- ntaxa(myphyseq_1)
  
# sequences after this stage
sample_sums_stage_1 <- sample_sums(myphyseq_1)

myphyseq_2 <- myphyseq_1

# further filtering and agglomeration (here, genus)
if(glom_level == "none"){
  # perform optional filtering
  if(removeKonly){
    myphyseq_2 <- subset_taxa(myphyseq_1, !is.na(phylum))
  }
} else {
  myphyseq_2 <- tax_glom(myphyseq_1, taxrank = glom_level)
  # NOTE FOR SELF: it would be safer to change taxa labels; physeq might retain the original labels
  # fix taxa_names
  taxa_names(myphyseq_2) <- myphyseq_2@tax_table[,glom_level]
}
sample_sums_stage_2 <- sample_sums(myphyseq_2)

ntaxa_post_glom <- ntaxa(myphyseq_2)

# final stage, filtering with prevab/permutation filter
myphyseq_3 <- myphyseq_2
# skipped if filterOTUs is set to F
# here I am filtering on prevalence and abundance, min prevalence 0.01, min abundance 0.005
if(filterOTUs){
  # prior to filtering must remove samples which have 0 sample_sums because
  # of taxonomic filtering and show message
  if(any(sample_sums(myphyseq_2)==0)){
    myphyseq_3 <- subset_samples(myphyseq_2, sample_sums(myphyseq_2)> 0)
    samples_removed <- names(which(sample_sums(myphyseq_2)==0))
    cat(red("\nsamples ",toString(samples_removed), "had 0 sequences left after taxonomic filtering and were removed\n"))
  }
  myfilteredlist <- filter_my_taxa(myphyseq_3, 
                                   name = "all_samples",
                           prevfilter = prev_filter,
                           PERFect_options = list(
                             method = "sim", # "sim" or "perm", perm takes longer
                             perf_alpha = 0.1, # this is the default
                             perf_k = 10000, # this is the number of permutation, default value
                             rollm = F, # this is the rolling mean argument
                             perf_alg = "fast", # algorithm for permutation filtering fast or full, full is slower
                             perf_hist = FALSE, # the hist option in PERFect_perm
                             vrb_out = TRUE # if extra output needs to be produced
                           ),
                           prevthreshold = prev_threshold,
                           passboth = pass_both,
                           abthreshold = ab_threshold,
                           filenm = "prevfilter",
                           gres = 300,
                           gtype = "tiff",
                           saveplot = T,
                           printplot = T,
                           outfolder = out_folder,
                           savepat = T
                           )
  # this generates a list which contains the phyloseq object
  myphyseq_3 <- myfilteredlist$phyobj
  ntaxa_post_prevab <- ntaxa(myphyseq_3)
  prevab_table <- myfilteredlist$prevabtable
  
  # TO DO, the table is sorted by decreasing prevalence need to provide an alternative my by a weighted sum
  # of prevalende and abundanceor something else
  
  filterOTU_flag <- T
}

# when assembling seq removal report need to keep into account if the length of vectors differ
sample_sums_stage_3 <- sample_sums(myphyseq_3)

if(exists("prevab_table")){
  prevab_table <- prevab_table |>
    mutate(label = .data[[glom_level]])
  # print(prevab_sort_option)
  sort_var = case_when(
    prevab_sort_option == "decr_prev" ~ prevab_table$relprev,
    prevab_sort_option == "decr_ab" ~ prevab_table$relAbundance,
    prevab_sort_option == "prevab" ~ 100*prevab_table$relprev*prevab_table$relAbundance,
    prevab_sort_option == "comp1" ~ pull(prevab_pc(prevab_df = prevab_table),Comp.1),
    .default = prevab_table$relprev
    )
  prevab_table$sort_var <- sort_var
  prevab_table <- arrange(prevab_table, desc(sort_var))
  top_taxa <- slice(prevab_table, 1:(maxOTUcats-1))
}

kable(
  select(top_taxa, label, phylum, class, relAbundance, relprev, sort_var),
  caption = str_c("Top ",(maxOTUcats-1), " bacterial genera in table olives and their food contact enviroments and materials in terms of prevalence and/or abundance", sep = ""), digits = dgts
  ) |>
  kable_classic()

```

Those genera do not completely cover dominating and subdominating taxa in all samples and aer more or less strongly affected by the unbalanced distribution of samples. Another option is showing, in a prevalence and abundance plot, all genera whose max abundance is > than a given threshold, say 0.01 and a given prevalence, say 0.1.  

```{r more_than_one_percemt, dpi = 96}

core_microbiota <- prevab_table |>
  dplyr::filter(max_rel_ab>0.01 & relprev >0.1) 
nrow(core_microbiota)
core_microbiota |>
  count(order) |>
  arrange(desc(n))

core_plot <- core_microbiota |>
  ggplot(mapping = aes(x = max_rel_ab, y = relprev)) +
  geom_point(aes(colour = order)) +
  geom_text_repel(aes(label = genus)) +
  labs(x = "max. rel. abundance",
       y= "rel. prevalence") +
  scale_color_brewer(type = "qual", palette = "Paired") +
  theme_bw()
print(core_plot)
if(savegraph) ggsave(core_plot, filename = file.path(out_folder, str_c(outfn_prefix,"core_all.",graphtype)), dpi = graphresolution)
```

Prevalence and abundance may, of course, differ in different groups of samples and is most certainly affected by the very unequal number of samples for different olive varieties. 

```{r after_filtering, echo = F}
samples_to_keep <- base::intersect(names(sample_sums_stage_0), names(sample_sums_stage_3))

seq_sums_df <- tibble(sample = samples_to_keep,
                         start = sample_sums_stage_0[samples_to_keep],
                         first_tax_filter = sample_sums_stage_1[samples_to_keep],
                         tax_glom = sample_sums_stage_2[samples_to_keep],
                         final_filt = sample_sums_stage_3[samples_to_keep])
seq_sums_df <- seq_sums_df |>
  mutate(prop_left = final_filt/start)

if(verbose_output) print(seq_sums_df, digits = dgts)

summary(ecdf(seq_sums_df$prop_left))

```
For some samples a large number of sequences was lost because of poor taxonomic assignment and/or a large number of non-bacterial sequences.  
There is also a risk that, in ordering the genera on the basis of the first principal component of prevalence and abundance, when choosing a subset of the most relevant taxa, some which are relevant only in a subset of samples will be discarderd. 

I will now repeat the filtering for the main groups of olives (Green Spanish style, Green Natural, Olives turning color Natural, Black Natural); environmental samples, which are found solely in the Nyons data sets, will be discarded.  

```{r prevab_by_group, dpi = 96, echo = F}

# final stage, filtering with prevab/permutation filter, for natural olives
ripe_trade_groups <- distinct(sample_data_B_extra, ripe_trade) |>
  pull()

# remove environment only
myphyseq_2_no_env <- subset_samples(myphyseq_2, !str_detect(sample_type, "contact"))
# nsamples(myphyseq_2_no_env)

# fix taxa_names
if(glom_level !="none"){
  taxa_names(myphyseq_2_no_env) <- myphyseq_2_no_env@tax_table[,glom_level]
}

# create container
prevab_by_ripe_trade <- vector(mode = "list", length = length(ripe_trade_groups))


for(i in seq_along(ripe_trade_groups)){
  myphyseq_3_temp <- subset_samples(myphyseq_2_no_env, ripe_trade == ripe_trade_groups[[i]])
  if(filterOTUs){
    # prior to filtering must remove samples which have 0 sample_sums because
    # of taxonomic filtering and show message
    if(any(sample_sums(myphyseq_3_temp)==0)){
      myphyseq_3_temp <- subset_samples(myphyseq_3_temp, sample_sums(myphyseq_3_temp)> 0)
    }
    myfilteredlist_temp <- filter_my_taxa(myphyseq_3_temp, 
                                          name = ripe_trade_groups[[i]],
                                          prevfilter = prev_filter,
                                          PERFect_options = list(
                                            method = "sim", # "sim" or "perm", perm takes longer
                                            perf_alpha = 0.1, # this is the default
                                            perf_k = 10000, # this is the number of permutation, default value
                                            rollm = F, # this is the rolling mean argument
                                            perf_alg = "fast", # algorithm for permutation filtering fast or full, full is slower
                                            perf_hist = FALSE, # the hist option in PERFect_perm
                                            vrb_out = TRUE # if extra output needs to be produced
                                          ),
                                          prevthreshold = prev_threshold,
                                          passboth = pass_both,
                                          abthreshold = ab_threshold,
                                          filenm = str_c("prevfilter", ripe_trade_groups[[i]]),
                                          gres = 300,
                                          gtype = "tiff",
                                          saveplot = T,
                                          printplot = T,
                                          outfolder = out_folder,
                                          savepat = T
    )
    prevab_by_ripe_trade[[i]] <- myfilteredlist_temp
    names(prevab_by_ripe_trade)[i] <- ripe_trade_groups[[i]]
  }
  
  prevab_table_temp <- myfilteredlist_temp$prevabtable  
  
  if(exists("prevab_table_temp")){
    # print(prevab_sort_option)
    sort_var_temp = case_when(
      prevab_sort_option == "decr_prev" ~ prevab_table_temp$relprev,
      prevab_sort_option == "decr_ab" ~ prevab_table_temp$relAbundance,
      prevab_sort_option == "prevab" ~ 100*prevab_table_temp$relprev*prevab_table_temp$relAbundance,
      prevab_sort_option == "comp1" ~ pull(prevab_pc(prevab_df = prevab_table_temp),Comp.1),
      .default = prevab_table_temp$relprev
    )
    prevab_table_temp$sort_var <- sort_var_temp
    prevab_table_temp <- arrange(prevab_table_temp, desc(sort_var))
    top_taxa_temp <- slice(prevab_table_temp, 1:(maxOTUcats-1))
    print(
      kable(
        select(top_taxa_temp, label, phylum, class, relAbundance, relprev, sort_var),
        caption = str_c("Top ", (maxOTUcats-1), 
                        " bacterial genera in table olives in terms of prevalence and/or abundance, ",
                        ripe_trade_groups[[i]], sep = ""), digits = dgts
      ))
    cat("\n")
  }
  
}

# clean up
rm(myphyseq_3_temp, myfilteredlist_temp, prevab_table_temp, i )
gc()


```


As expected, the pattern of most abundant and prevalent genera is different in each group.  
Therefore, to avoid the possibility that a taxon that is important in a group but not in another is discarded, we created a list of taxa to be retained which was the union of taxa retained in each group.

```{r modified_filter}

prevab_all <- map(prevab_by_ripe_trade, \(x) pluck(x,"prevabtable")) |>
  list_rbind(names_to = "ripe_trade") |>
  dplyr::filter(pass_filters == "T")

taxa_to_keep <- distinct(prevab_all, label) |> pull()

myfiltphyseq <- subset_taxa(myphyseq_2, taxa_names(myphyseq_2) %in% taxa_to_keep)

```


# The most abundant genera.

I will now create box plots and bar plots by group, using olive or brine samples only, with the 30 most prevalent and abundant genera. I will use all sample categories.  
It might also be interesting to show groups, like _Lactobacillaceae_, halophilic microorganisms (halophiles + HALAB) and common spoilage microorganisms.

```{r top_taxa_box_plot, dpi = 96}
# get the taxa in the core microbiota
core_label <- core_microbiota$label
# create a phyloseq with relative abundances, preserving the OTUs which were not filtered
myphyseq_2_relab <- transform_sample_counts(myphyseq_2, \(x) x/sum(x))
# melt and select only essential variables
myphyseq_2_melt <- psmelt(myphyseq_2_relab) |>
  select(OTU:studyId,s_type,L1:spoil_ferm, olive_variety:inoculation_fortification,
         olive_var_short, ripe_trade, class,family,genus)
# create stage variable
myphyseq_2_melt <- myphyseq_2_melt |> mutate(stage_category = case_when(
    nature == "Raw" ~ 1,
    nature == "Intermediate" ~ 2,
    nature == "Finished" ~ 3,
    .default = 0
  ))

# filter for the top 30
myphyseq_2_melt_core <- myphyseq_2_melt |> dplyr::filter(OTU %in% core_label)

# need a reasonable ordering for x axis
order_x <- myphyseq_2_melt_core |>
  select(OTU, family) |>
  distinct() |>
  arrange(family, OTU) |>
  pull(OTU)

boxplot_core <- myphyseq_2_melt_core |>
  dplyr::filter(!str_detect(sample_type, "contact")) |>
  mutate(OTU = factor(OTU, levels = order_x)) |>
  ggplot(mapping = aes(x = OTU, y = log10(Abundance))) +
  geom_jitter(aes(color = olive_var_short), alpha = I(0.7)) +
  facet_wrap(~ripe_trade) +
  geom_boxplot(alpha=I(0.5)) +
  labs(color = "olive variety", y = "log10(rel. abund.)") +
  scale_color_brewer(type = "qual", palette = "Paired") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, face = "italic"))
boxplot_core
if(savegraph) ggsave(boxplot_core, filename = file.path(out_folder,str_c(outfn_prefix, "boxplotcore.",graphtype)), dpi = graphresolution, width = 9)
```

This plot gives emphasis to variability, but it is strongly affected by unequal sampling levels, especially for black and turning color natural categories (almost all samples are from Nyons olives in the first case).

```{r boxplot_LAB, dpi = 96}
# try with subgroups
# LAB, only those passing filter
LAB_pass_filter <- prevab_table |>
  dplyr::filter(pass_filters == "T" & (family %in% c("Lactobacillaceae","Streptococcaceae", "Carnobacteriaceae", "Enterococcaceae"))) |>
  select(label, family, genus) |>
  arrange(family, genus) |>
  pull(label)

boxplot_LAB <- myphyseq_2_melt |>
  dplyr::filter(OTU %in% LAB_pass_filter) |>
  mutate(OTU=factor(OTU, levels = LAB_pass_filter)) |> 
  ggplot(mapping = aes(x = OTU, y = log10(Abundance))) +
  geom_jitter(aes(color = olive_var_short, shape = inoculation_fortification), alpha = I(0.7)) +
  facet_wrap(~ripe_trade) +
  geom_boxplot(alpha=I(0.5)) +
  labs(color = "olive variety", 
       y = "log10(rel. abund.)",
       shape = "Inoculation/fortification") +
  scale_color_brewer(type = "qual", palette = "Paired") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, face = "italic"))
boxplot_LAB
if(savegraph) ggsave(boxplot_LAB,
                     filename = file.path(out_folder,str_c(outfn_prefix, "boxplotLAB.",graphtype)),
                     dpi = graphresolution, width = 9, height = 7)

```

The LAB plot is simpler to interpret and also shows some minor or more rare genera. HALAB are definitely more prevalent and abundant in Gee Spanish style olives, with some exceptions.

```{r boxplot_halo, dpi = 96}
# other halophilic genera
halo_noHALAB <- c("Aliidiomarina","Celerinatantimonas", "Chromohalobacter", "Halomonas", 
                  "Idiomarina","Marinobacter", "Marinomonas",  "Salinicola","Staphylococcus", "Vibrio")
Halo_pass_filter <- prevab_table |>
  dplyr::filter(pass_filters == "T" & (genus %in% halo_noHALAB)) |>
  select(label, family, genus) |>
  arrange(family, genus) |>
  pull(label)

boxplot_halo <- myphyseq_2_melt |>
  dplyr::filter((OTU %in% Halo_pass_filter)) |>
  mutate(OTU=factor(OTU, levels = Halo_pass_filter)) |> 
  ggplot(mapping = aes(x = OTU, y = log10(Abundance))) +
  geom_jitter(aes(color = olive_var_short, shape = inoculation_fortification), alpha = I(0.7)) +
  facet_wrap(~ripe_trade) +
  geom_boxplot(alpha=I(0.5)) +
  labs(color = "olive variety", y = "log10(rel. abund.)", shape = "inoculation/fortification") +
  scale_color_brewer(type = "qual", palette = "Paired") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, face = "italic"))
boxplot_halo
if(savegraph) ggsave(boxplot_halo, filename = file.path(out_folder,str_c(outfn_prefix, "boxplotHALO.",graphtype)), dpi = graphresolution)


```
The prevalence and abundance is very clearly affected by olive ripeness and trade preparation but also by variety.  
Must check in Penland et al. if they have proven that salt is a source.  

## Occurrence of genera which may include pathogenic bacteria in table olives.  

Modern HTS return very large numbers of tags per sample. This, in theory, might give us the opportunity to use metataxonomic datasets to assess the presence of genomic DNA of pathogenic microorganisms. In FoodMicrobionet we do taxonomic assignment at the species level, whenever possible (sequences of good quality and sufficient length, successful merging of paired ends are a prerequisite) but, due to the heterogeneity of data and the difficulties of taxonomic assignment with short reads, species assignment is, at its best, dubious.
The following analysis provides some data on the occurrence of sequences belonging to selected genera which include pathogenic groups (the list is taken from the [Bad Bug Book](https://www.fda.gov/food/foodborne-pathogens/bad-bug-book-second-edition)). Some genera (_Arcobacter_, _Aeromonas_, _Bacillus_, _Cronobacter_, _Clostridium_, _Enterococcus_, _Escherichia_, _Mycobacterium_,  _Staphylococcus_, _Streptococcus_, _Vibrio_) which include mostly non-pathogenic species or strains (some of which are, indeed, beneficial in food fermentations) are excluded.

```{r pathogenic_bacteria}
patho_species <- c("Aeromonas hydrophila", "Arcobacter butzleri",  "Bacillus cereus",
                   "Brucella abortus", "Brucella melitensis",
                   "Campylobacter jejuni", "Campylobacter coli", "Campylobacter fetus",
                   "Clostridium botulinum", "Clostridium perfringes", 
                   "Coxiella burnetii", "Cronobacter sakazakii", 
                   "Listeria monocytogenes",
                   "Mycobacterium bovis", "Plesiomonas shigelloides",
                   "Salmonella enterica", "Shigella dysenteriae", "Shigella sonnei", 
                   "Shigella boydii", "Staphylococcus aureus", "Staphylococcus hyicus",
                   "Streptococcus pyogenes", "Streptococcus zooepidemicus", 
                   "Vibrio cholerae", "Vibrio parahaemolyticus",
                   "Vibrio vulnificus", "Yersinia enterocolitica")
# are these species found in the taxa table?
any(patho_species %in% taxa_B[,7])

patho_genera <- c("Brucella", "Campylobacter", "Listeria", "Plesiomonas", "Shigella", 
                  "Salmonella", "Yersinia")
# which of these are in the original taxa table?
patho_in_table <- patho_genera[patho_genera %in% taxa_B[,6]]
patho_in_table
# build a box plot from filtered phyloseq
# use a melted phyloseq
patho_box_plot <- myphyseq_2_melt |>
  dplyr::filter((OTU %in% patho_in_table)) |>
  mutate(OTU=factor(OTU, levels = patho_in_table)) |> 
  ggplot(mapping = aes(x = OTU, y = log10(Abundance))) +
  geom_jitter(aes(color = olive_var_short, shape = inoculation_fortification), alpha = I(0.7)) +
  facet_wrap(~ripe_trade) +
  geom_boxplot(alpha=I(0.5)) +
  labs(color = "olive variety", y = "log10(rel. abund.)") +
  scale_color_brewer(type = "qual", palette = "Paired") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, face = "italic"))
patho_box_plot 
if(savegraph) ggsave(patho_box_plot , 
                     filename = file.path(out_folder, 
                                          str_c(outfn_prefix, "boxplotpatho.",graphtype)), 
                     dpi = graphresolution)
# figures on prevalence
patho_genera_df <-  myphyseq_2_melt |>
  dplyr::filter((OTU %in% patho_in_table)) |>
  mutate(OTU=factor(OTU, levels = patho_in_table))

prevab_patho <- prevab_table |> 
  dplyr::filter(label %in% patho_genera) |>
  select(label, Prevalence, TotalAbundance, max_rel_ab, relAbundance, relprev)

kable(prevab_patho, caption = "Table x. Data on prevalence and abundance of genera including pathogenic microorganisma in the METAolive dataset") |>
  kableExtra::kable_classic()

```


## A bar plot.  

Finally, I will try to depict the composition of the dominant microbiota different olive varieties at different stages of fermentaion. The data are very unbalanced and, especially for very long fermentations (Nyons, Itrana), the stage 2 may pool together samples which are very far apart in time.  

```{r top_taxa_barplot, echo = F, dpi = 72}

# use the core_microbiota subset as starting point

# get the OTU table
OTU_mat_f <- as(otu_table(myfiltphyseq), "matrix")
if (myfiltphyseq@otu_table@taxa_are_rows) {
    OTU_mat_f <- t(OTU_mat_f)
}  
# calculate OTU sums by sample
OTU_mat_f_sums <- rowSums(OTU_mat_f)
# extract top_taxa
OTU_mat_f_core <- OTU_mat_f[,core_microbiota$label]
coresums <- rowSums(OTU_mat_f_core)
Other <- OTU_mat_f_sums-coresums
OTU_mat_f_coreo <- cbind(OTU_mat_f_core, Other)
# make a tibble
OTU_mat_f_coreo <- as_tibble(OTU_mat_f_coreo, rownames="Run")
# pivot longer and remove 0s
OTU_tibble_long <- pivot_longer(OTU_mat_f_coreo, cols = 2:ncol(OTU_mat_f_coreo),
                                names_to = "label", values_to = "nseq") |>
  dplyr::filter(nseq>0) 

# join sample metadata
OTU_tibble_long <- left_join(OTU_tibble_long, sample_data_B_2)

# join taxa metadata

OTU_tibble_long <- left_join(OTU_tibble_long,
                             select(core_microbiota, label, domain:family))
if(glom_level != "none"){
  var <- "label"
  OTU_tibble_long <- mutate(OTU_tibble_long, 
                            tax_name = if_else(label !="Other", .data[[var]], "Other")) %>%
    mutate(stage_category = case_when(
    nature == "Raw" ~ 1,
    nature == "Intermediate" ~ 2,
    nature == "Finished" ~ 3,
    .default = 0
  ))
  
}

# OK now let's group samples (I am using this method to create a Shiny app later)
grouping_vars <- c("ripe_trade", "group_short", "stage_category", "tax_name")

OTU_tibble_sums <- OTU_tibble_long %>%
  summarise(seq_sums = sum(nseq), .by = all_of(grouping_vars))
# make the taxa labels a factor, reordered on the basis of family
order_sums <- core_microbiota |>
  select(label, family) |>
  distinct() |>
  arrange(family, label) |>
  pull(label)
order_sums %in% unique(OTU_tibble_sums$tax_name)

OTU_tibble_sums <- OTU_tibble_sums |>
  mutate(tax_name = factor(tax_name, levels = c(order_sums, "Other")))

# make a qualitative color palette
n_taxa <- nrow(core_microbiota)+1
rpalette <- distinctColorPalette(n_taxa)
names(rpalette) <- NULL


if(maxOTUcats<=11){
  top_bar_plot <- OTU_tibble_sums |>
    ggplot(mapping = aes(x = group_short, y  = seq_sums, fill = tax_name)) +
    facet_grid(stage_category ~ str_wrap(ripe_trade,20), scales = "free_x") +
    geom_col(position = "fill") +
    scale_fill_brewer(type = "qual", palette = "Paired") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
} else {
    top_bar_plot <- OTU_tibble_sums |>
    ggplot(mapping = aes(x = group_short, y  = seq_sums, fill = tax_name)) +
    facet_grid(stage_category ~ str_wrap(ripe_trade,20), scales = "free_x") +
    geom_col(position = "fill") +
    scale_fill_manual(values = rpalette[1:n_taxa]) +
    labs(fill = NULL) +
    guides(fill = guide_legend(ncol = 6)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          legend.position = "bottom")
}
top_bar_plot
# this should be conditional to savegraph
if(savegraph) ggsave(top_bar_plot, filename = file.path(out_folder, str_c("totaxabchart", gfileext, sep="")), 
       dpi = graphresolution, width = 9, height = graphsizein,
       units = "in")
```

This display gives emphasis to a comparison between different trade preparations for the same category at the same time stage. Of course a different emphasis could be used. 
Apparently, even selecting a high number of dominating genera (31) we are unable to cover all the diversity in table olives.
It is worth noticing that the taxonomic group "Other" may be taken, generally speaking, as a measure of diversity: the larger the section, the larger the diversity. This is clearly visible in the raw fruits. The "2" category pools all the samples during fermentation, some of which can be several months apart.  
In several, but not all cases, the composition of the bacterial microbiota of brine and fruit and, when available, olive contact surfaces and material, differ within the same variety.     
In addition, the display is clearly dominated by the varieties for which more data are available. Note also that here spoiled or inoculated samples are pooled with the others.   

## A multidimensional scaling, fast and dirty.  

Here I will not insist much on beta diversity: the choice of graph and analyses should be guided by scientific questions. I will only perform a fast and dirty non-metric multidimensional scaling on the Bray-Curtis distance matrix and enrich the plot with metadata. I think using the bipartite package might also be interesting to detect modules and to calculate statistics on nodes, but a reasonable grouping of samples must be found.  

```{r MDS, echo = F, }

myord <- ordinate(myphyseq_3, method = "NMDS", distance = "bray", trymax = 30)

# NOTE for self if I ever need explained variance # https://cran.r-project.org/web/packages/vegan/vignettes/FAQ-vegan.html
# extract coordinates of samples, add metadata
sample_scores <- as_tibble(myord$points, rownames = "Run") |>
  left_join(sample_data_B_2) |>
mutate(stage_category = case_when(
    nature == "Raw" ~ 1,
    nature == "Intermediate" ~ 2,
    nature == "Finished" ~ 3,
    .default = 0
  ))

sample_scores |>
  ggplot(mapping = aes(x= MDS1, y = MDS2)) +
  geom_point(mapping=aes(shape = sample_type, colour = inoculation_fortification, 
                         fill = ripe_trade, 
                         size = stage_category), alpha = I(0.5), stroke = 1) +
  geom_text(mapping = aes(label = olive_var_short), alpha = I(0.3), size = 2) +
  labs(fill = NULL, colour = NULL, shape = NULL, size = "stage") +
  scale_shape_manual(values=c(21,24,22,23)) +
  scale_colour_manual(values = c("blue","grey")) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  guides(fill = guide_legend("trade prep", override.aes = list(shape = 21))) +
  theme_bw()
if(savegraph) ggsave(filename = file.path(out_folder, str_c("NMDSsamples", gfileext, sep="")), 
       dpi = graphresolution, width = graphsizein, height = graphsizein,
       units = "in")


```

The ordination is far from ideal, with final stress oscillating arounf 0.205. There is still the problem of showing spoiled samples...
Anyway, olive variety seems more important than other factors and the addition of starters does not necessarily make a big difference.  
I should also consider not using stage 0 for brine, to make comparison with fruits of the same stage possible. This can possibly be done using the fermentation time rather than the nature field. I think it can be done programmatically with some effort. It is also possible that there are too many features in the same graph.  

I am now trying to show spoilage rather than inoculation.

```{r NMDS_spoilage, dpi = 96}
sample_scores |>
  ggplot(mapping = aes(x= MDS1, y = MDS2)) +
  geom_point(mapping=aes(shape = sample_type, colour = spoilage, 
                         fill = ripe_trade, 
                         size = stage_category), alpha = I(0.5), stroke = 1) +
  geom_text(mapping = aes(label = olive_var_short), alpha = I(0.3), size = 2) +
  labs(fill = NULL, colour = NULL, shape = NULL, size = "stage") +
  scale_shape_manual(values=c(21,24,22,23)) +
  scale_colour_manual(values = c("blue","red", "green")) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  guides(fill = guide_legend("trade prep", override.aes = list(shape = 21))) +
  theme_bw()
if(savegraph) ggsave(filename = file.path(out_folder, str_c("NMDSsamples_spoilage", gfileext, sep="")), 
       dpi = graphresolution, width = graphsizein, height = graphsizein,
       units = "in")
```


```{r species}
genus_tibble <- as(tax_table(myphyseq_3), "matrix") |>
  as_tibble(rownames = "tlabel")

taxa_scores <- as_tibble(myord$species, rownames = "tlabel") |>
  left_join(genus_tibble)

taxa_scores |> 
  dplyr::filter(genus %in% top_taxa$genus) |>
  ggplot(mapping=aes(x=MDS1, y = MDS2, color = class)) +
  geom_point() +
  geom_text(mapping= aes(label=genus), show_legend = F)

```
## More things to try.  

I could try a constrained ordination, but frankly the dataset is too unbalances and the ordination would probably be driven by the two varieties which have the most samples (Nyons and Itrana) which are also confounded with the variable ripe_trade.  

# References.  

Here are the references for the studies retrieved from FMBN.  

```{r references, echo = F}
references <- tibble(studyId = rep("Foodmicrobionet",4),
                     ref_complete = c(
                       paste0("Parente, E., Cocolin, L., De Filippis, F., Zotta, T., ",
                              "Ferrocino, I., O’Sullivan, O., Neviani, E., De Angelis, ",
                              "M., Cotter, P. D., Ercolini, D. 2016. FoodMicrobionet: a ",
                              "database for the visualisation and exploration of food ",
                              "bacterial communities based on network analysis. Int. J. ",
                              "Food Microbiol. 219: 28-37.", sep =""),
                       paste0("De Filippis, F., Parente, E., Zotta, T., Ercolini, D. 2018.", 
                              "A comparison of bioinformatic approaches for 16S rRNA gene ",
                              "profiling of food bacterial microbiota. Int. J. Food Microbiol. 265:9-17.",
                              sep = ""),
                       paste0("Parente, E., De Filippis, F., Ercolini, D., Ricciardi, A., Zotta, T., 2019. ", 
                              "Advancing integration of data on food microbiome studies: FoodMicrobionet ",
                              "3.1, a major upgrade of the FoodMicrobionet database. Int. J. Food Microbiol. 305:108249.",
                              sep = ""),
                       paste0("Parente, E., Zotta, T., Ricciardi, A., 2022. ", 
                              "FoodMicrobionet v4: A large, integrated, open and transparent database for food bacterial communities. ",
                              "Int J Food Microbiol 372, 109696.",
                              sep = "")),
                     DOI = c("10.1016/j.ijfoodmicro.2015.12.001",
                             "10.1016/j.ijfoodmicro.2017.10.028",
                             "10.1016/j.ijfoodmicro.2019.108249",
                             "10.1016/j.ijfoodmicro.2022.109696")
)
references <- bind_rows(references,
                        select(studies_B, studyId, ref_complete, DOI))
kable(references, style = "html")
```



# Credits and copyright.

This is version 0.3 of the script, 5/1.  

The code for loading packages is derived from [this post](http://tinyurl.com/jjwyzph).  
The code for making qualitative random palette is derived from [this article](http://tinyurl.com/hawqufy).  
The code for heat maps is modified from an article on the [Molecular ecologist blog](http://tinyurl.com/hxxbmvz).

Assume that the code in this document is overall under MIT licence

Permission is hereby granted, free of charge, to any person obtaining a copy of this software and associated documentation files (the "Software"), to deal in the Software without restriction, including without limitation the rights to use, copy, modify, merge, publish, distribute, sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is furnished to do so, subject to the following conditions: THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.


## Citations for R packages used in this document.  

```{r r_packages, echo=F, include = T}
all_packages <- c("base", .cran_packages, .bioc_packages)
print(map(all_packages, citation, auto = T), bibtex = F)

```

## Session report 
This script has been tested with R 4.3.3, x86 build, on a MacBook Pro 14 inch, 2021, with chip Apple M1 Pro and 16 GB RAM and on a MacStudio 2023 with Apple M2 Max chip, both running MacOs Sonoma 12.3.1


```{r not_run, eval = F}
sessionInfo()
```

## Acknowledgements 

This work was was carried out within the PRIN 2022 Project METAOlive 2022NN28ZZ and received funding from the European Union Next-GenerationEU, CUP C53D23005460006 (PIANO NAZIONALE DI RIPRESA E RESILIENZA (PNRR) – MISSIONE 4 COMPONENTE 2,  INVESTIMENTO 1.4 – D.D. 1048 14/07/2023). This script and its contents reflects only the authors’ views and opinions,  neither the European Union nor the European Commission can be considered  responsible for them.

